Counterfactual Training: Teaching Models Plausible and Actionable Explanations

### 2025-03-13 paper/preprint/paper.tex
Review abstract and introduction
final table done

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}

@@

\section{Detailed Results}\label{detailed-results}

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]

@@

\subsection{Grid Searches}\label{sec-app-grid}

@@

\subsubsection{Evaluation Details}\label{evaluation-details}

@@

\paragraph{Predictive Performance}\label{predictive-performance}

@@

\paragraph{Counterfactual Outcomes}\label{sec-app-grid-hl}

Overall, we find that Counterfactual Training (CT) achieves it key
objectives consistently across all hyperparameter settings and also
broadly across datasets: plausibility is improved by up to
\textasciitilde60 percentage points (ppts) for the \emph{Circles} data
(e.g. Figure~\ref{fig-grid-gen_params-plaus-circles}),
\textasciitilde25-30ppts for the \emph{Moons} data (e.g.
Figure~\ref{fig-grid-gen_params-plaus-moons}) and
\textasciitilde10-20ppts for the \emph{Linearly Separable} data (e.g.
Figure~\ref{fig-grid-gen_params-plaus-lin_sep}). At the same time, the
average costs of faithful counterfactuals are reduced in many cases by
around \textasciitilde20-25ppts for \emph{Circles} (e.g.
Figure~\ref{fig-grid-gen_params-cost-circles}) and up to
\textasciitilde50ppts for \emph{Moons} (e.g.
Figure~\ref{fig-grid-gen_params-cost-moons}). For the \emph{Linearly
Separable} data, costs are generally increased although typically by
less than 10ppts (e.g. Figure~\ref{fig-grid-gen_params-cost-lin_sep}),
which reflects a common tradeoff between costs and plausibility
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).

@@

\subsubsection{Generator Parameters}\label{sec-app-grid-gen}

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]

@@

\paragraph{Accuracy}\label{accuracy}

@@

  \toprule
  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
  \endfirsthead
  \toprule
  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
  \endhead
  \bottomrule
  \multicolumn{5}{r}{Continuing table below.}\\
  \bottomrule
  \endfoot
  \endlastfoot
  Circles & Accuracy & Full & 0.997 & 0.00309 \\
  Circles & Accuracy & Vanilla & 0.998 & 0.000557 \\
  Circles & F1-score & Full & 0.997 & 0.00309 \\
  Circles & F1-score & Vanilla & 0.998 & 0.000558 \\
  Lin Sep & Accuracy & Full & 0.999 & 0.00201 \\
  Lin Sep & Accuracy & Vanilla & 1 & 0 \\
  Lin Sep & F1-score & Full & 0.999 & 0.00201 \\
  Lin Sep & F1-score & Vanilla & 1 & 0 \\
  Moons & Accuracy & Full & 0.999 & 0.000696 \\
  Moons & Accuracy & Vanilla & 1 & 0.00111 \\
  Moons & F1-score & Full & 0.999 & 0.000696 \\
  Moons & F1-score & Vanilla & 1 & 0.00111 \\
  Over & Accuracy & Full & 0.915 & 0.00477 \\
  Over & Accuracy & Vanilla & 0.917 & 0.00123 \\
  Over & F1-score & Full & 0.915 & 0.00478 \\
  Over & F1-score & Vanilla & 0.917 & 0.00124 \\\bottomrule

@@

\paragraph{Plausibility}\label{plausibility-2}

@@

\paragraph{Cost}\label{cost}

@@

\subsubsection{Penalty Strengths}\label{sec-app-grid-pen}

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]

@@

\paragraph{Accuracy}\label{accuracy-1}

@@

  \toprule
  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
  \endfirsthead
  \toprule
  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
  \endhead
  \bottomrule
  \multicolumn{5}{r}{Continuing table below.}\\
  \bottomrule
  \endfoot
  \endlastfoot
  Circles & Accuracy & Full & 0.994 & 0.0144 \\
  Circles & Accuracy & Vanilla & 0.998 & 0.000875 \\
  Circles & F1-score & Full & 0.994 & 0.0145 \\
  Circles & F1-score & Vanilla & 0.998 & 0.000875 \\
  Lin Sep & Accuracy & Full & 0.998 & 0.00772 \\
  Lin Sep & Accuracy & Vanilla & 1 & 0 \\
  Lin Sep & F1-score & Full & 0.998 & 0.00773 \\
  Lin Sep & F1-score & Vanilla & 1 & 0 \\
  Moons & Accuracy & Full & 0.987 & 0.0351 \\
  Moons & Accuracy & Vanilla & 0.998 & 0.0101 \\
  Moons & F1-score & Full & 0.987 & 0.0352 \\
  Moons & F1-score & Vanilla & 0.998 & 0.0102 \\
  Over & Accuracy & Full & 0.911 & 0.0217 \\
  Over & Accuracy & Vanilla & 0.916 & 0.00236 \\
  Over & F1-score & Full & 0.911 & 0.0219 \\
  Over & F1-score & Vanilla & 0.916 & 0.00236 \\\bottomrule

@@

\paragraph{Plausibility}\label{plausibility-3}

@@

\paragraph{Cost}\label{cost-1}

@@

\subsubsection{Other Parameters}\label{sec-app-grid-train}

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-train}: Training Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]

@@

\paragraph{Accuracy}\label{accuracy-2}

@@

  \toprule
  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
  \endfirsthead
  \toprule
  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
  \endhead
  \bottomrule
  \multicolumn{5}{r}{Continuing table below.}\\
  \bottomrule
  \endfoot
  \endlastfoot
  Circles & Accuracy & Full & 0.995 & 0.00431 \\
  Circles & Accuracy & Vanilla & 0.998 & 0.000566 \\
  Circles & F1-score & Full & 0.995 & 0.00432 \\
  Circles & F1-score & Vanilla & 0.998 & 0.000566 \\
  Lin Sep & Accuracy & Full & 0.999 & 0.00231 \\
  Lin Sep & Accuracy & Vanilla & 1 & 0 \\
  Lin Sep & F1-score & Full & 0.999 & 0.00231 \\
  Lin Sep & F1-score & Vanilla & 1 & 0 \\
  Moons & Accuracy & Full & 0.996 & 0.0136 \\
  Moons & Accuracy & Vanilla & 0.988 & 0.022 \\
  Moons & F1-score & Full & 0.996 & 0.0136 \\
  Moons & F1-score & Vanilla & 0.988 & 0.022 \\
  Over & Accuracy & Full & 0.914 & 0.00563 \\
  Over & Accuracy & Vanilla & 0.918 & 0.00116 \\
  Over & F1-score & Full & 0.914 & 0.0057 \\
  Over & F1-score & Vanilla & 0.918 & 0.00116 \\\bottomrule

@@

\paragraph{Plausibility}\label{plausibility-4}

@@

\paragraph{Cost}\label{cost-2}

@@

\subsection{Hyperparameter Tuning}\label{sec-app-tune}

@@

Finally, to increase the proportion of mature counterfactuals for some
datasets, we have also investigated the effect on the learning rate
\(\eta\) for the counterfactual search and even smaller regularization
strengths for a fixed decision threshold of 0.5
(Figure~\ref{fig-tune_lr-plaus-adult} to
Figure~\ref{fig-tune_lr-plaus-over}). For the given low decision
threshold, we find that the learning rate has no discernable impact on
the proportion of mature counterfactuals
(Figure~\ref{fig-tune_lr-mat-adult} to
Figure~\ref{fig-tune_lr-mat-over}). We do notice, however, that the
results for MNIST are much improved when using a low value
\(\lambda_{\text{reg}}\), the strength for the engery regularization:
plausibility is increased by up to \textasciitilde10ppt
(Figure~\ref{fig-tune_lr-plaus-mnist}) and the proportion of mature
counterfactuals reaches 100\%.

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-warning-color-frame, title={Package Version (Reproducibility)}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-warning-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]

@@

\subsubsection{Key Parameters}\label{sec-app-tune-key}

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-train}: Training Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]

@@

\paragraph{Plausibility}\label{plausibility-5}

@@

\paragraph{Proportion of Mature CE}\label{proportion-of-mature-ce}

@@

\subsubsection{Learning Rate}\label{sec-app-tune-lr}

@@

\paragraph{Plausibility}\label{plausibility-6}

@@

\paragraph{Proportion of Mature CE}\label{proportion-of-mature-ce-1}

>>>>

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}

@@

\section{Details on Main Experiments}\label{details-on-main-experiments}

\subsection{Final Hyperparameters}\label{final-hyperparameters}

As discussed Section~\ref{sec-experiments}, CT is sensitive to certain
hyperparameter choices. We study the effect of many hyperparameters
extensively in Section~\ref{sec-app-grid}. For the main results, we tune
a small set of key hyperparameters (Section~\ref{sec-app-tune}). The
final choices for the main results are presented for each data set in
Table~\ref{tbl-final-params} along with training, test and batch sizes.

\begin{longtable}{ccccccc}

\caption{\label{tbl-final-params}Final hyperparameters used for the main
results for the different datasets.}

  \toprule
  \textbf{Data} & \textbf{No. Train} & \textbf{No. Test} & \textbf{Batchsize} & \textbf{Decision Threshold} & \textbf{No. Counterfactuals} & \textbf{$\lambda_{\text{reg}}$} \\\midrule
  \endfirsthead
  \toprule
  \textbf{Data} & \textbf{No. Train} & \textbf{No. Test} & \textbf{Batchsize} & \textbf{Decision Threshold} & \textbf{No. Counterfactuals} & \textbf{$\lambda_{\text{reg}}$} \\\midrule
  \endhead
  \bottomrule
  \multicolumn{7}{r}{Continuing table below.}\\
  \bottomrule
  \endfoot
  \endlastfoot
  Adult & $2.6 \cdot 10^{4}$ & $5.01 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & 0.75 & $5 \cdot 10^{3}$ & 0.25 \\
  California Housing & $1.65 \cdot 10^{4}$ & $3.1 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & 0.5 & $5 \cdot 10^{3}$ & 0.25 \\
  Circles & $3.6 \cdot 10^{3}$ & 600 & 30 & 0.5 & $1 \cdot 10^{3}$ & 0.5 \\
  Credit & $1.06 \cdot 10^{4}$ & $1.92 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & 0.5 & $5 \cdot 10^{3}$ & 0.25 \\
  GMSC & $1.34 \cdot 10^{4}$ & $2.47 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & 0.5 & $5 \cdot 10^{3}$ & 0.5 \\
  Linearly Separable & $3.6 \cdot 10^{3}$ & 600 & 30 & 0.5 & $1 \cdot 10^{3}$ & 0.25 \\
  Moons & $3.6 \cdot 10^{3}$ & 600 & 30 & 0.9 & $1 \cdot 10^{3}$ & 0.25 \\
  Overlapping & $3.6 \cdot 10^{3}$ & 600 & 30 & 0.5 & $1 \cdot 10^{3}$ & 0.25 \\\bottomrule

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]

@@

\section{Grid Searches}\label{sec-app-grid}

@@

\subsection{Evaluation Details}\label{evaluation-details}

@@

\subsubsection{Predictive Performance}\label{predictive-performance}

@@

\subsubsection{Counterfactual Outcomes}\label{sec-app-grid-hl}

Overall, we find that Counterfactual Training (CT) achieves it key
objectives consistently across all hyperparameter settings and also
broadly across datasets: plausibility is improved by up to
\textasciitilde60 percent (\%) for the \emph{Circles} data (e.g.
Figure~\ref{fig-grid-gen_params-plaus-circles}), \textasciitilde25-30\%
for the \emph{Moons} data (e.g.
Figure~\ref{fig-grid-gen_params-plaus-moons}) and \textasciitilde10-20\%
for the \emph{Linearly Separable} data (e.g.
Figure~\ref{fig-grid-gen_params-plaus-lin_sep}). At the same time, the
average costs of faithful counterfactuals are reduced in many cases by
around \textasciitilde20-25\% for \emph{Circles} (e.g.
Figure~\ref{fig-grid-gen_params-cost-circles}) and up to
\textasciitilde50\% for \emph{Moons} (e.g.
Figure~\ref{fig-grid-gen_params-cost-moons}). For the \emph{Linearly
Separable} data, costs are generally increased although typically by
less than 10\% (e.g. Figure~\ref{fig-grid-gen_params-cost-lin_sep}),
which reflects a common tradeoff between costs and plausibility
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).

@@

\subsection{Generator Parameters}\label{sec-app-grid-gen}

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]

@@

\subsubsection{Accuracy}\label{accuracy}

@@

  \toprule
  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
  \endfirsthead
  \toprule
  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
  \endhead
  \bottomrule
  \multicolumn{5}{r}{Continuing table below.}\\
  \bottomrule
  \endfoot
  \endlastfoot
  Circles & Accuracy & Full & 0.997 & 0.00309 \\
  Circles & Accuracy & Vanilla & 0.998 & 0.000557 \\
  Circles & F1-score & Full & 0.997 & 0.00309 \\
  Circles & F1-score & Vanilla & 0.998 & 0.000558 \\
  Linearly Separable & Accuracy & Full & 0.999 & 0.00201 \\
  Linearly Separable & Accuracy & Vanilla & 1 & 0 \\
  Linearly Separable & F1-score & Full & 0.999 & 0.00201 \\
  Linearly Separable & F1-score & Vanilla & 1 & 0 \\
  Moons & Accuracy & Full & 0.999 & 0.000696 \\
  Moons & Accuracy & Vanilla & 1 & 0.00111 \\
  Moons & F1-score & Full & 0.999 & 0.000696 \\
  Moons & F1-score & Vanilla & 1 & 0.00111 \\
  Overlapping & Accuracy & Full & 0.915 & 0.00477 \\
  Overlapping & Accuracy & Vanilla & 0.917 & 0.00123 \\
  Overlapping & F1-score & Full & 0.915 & 0.00478 \\
  Overlapping & F1-score & Vanilla & 0.917 & 0.00124 \\\bottomrule

@@

\subsubsection{Plausibility}\label{plausibility-2}

@@

\subsubsection{Cost}\label{cost}

@@

\subsection{Penalty Strengths}\label{sec-app-grid-pen}

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]

@@

\subsubsection{Accuracy}\label{accuracy-1}

@@

  \toprule
  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
  \endfirsthead
  \toprule
  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
  \endhead
  \bottomrule
  \multicolumn{5}{r}{Continuing table below.}\\
  \bottomrule
  \endfoot
  \endlastfoot
  Circles & Accuracy & Full & 0.994 & 0.0144 \\
  Circles & Accuracy & Vanilla & 0.998 & 0.000875 \\
  Circles & F1-score & Full & 0.994 & 0.0145 \\
  Circles & F1-score & Vanilla & 0.998 & 0.000875 \\
  Linearly Separable & Accuracy & Full & 0.998 & 0.00772 \\
  Linearly Separable & Accuracy & Vanilla & 1 & 0 \\
  Linearly Separable & F1-score & Full & 0.998 & 0.00773 \\
  Linearly Separable & F1-score & Vanilla & 1 & 0 \\
  Moons & Accuracy & Full & 0.987 & 0.0351 \\
  Moons & Accuracy & Vanilla & 0.998 & 0.0101 \\
  Moons & F1-score & Full & 0.987 & 0.0352 \\
  Moons & F1-score & Vanilla & 0.998 & 0.0102 \\
  Overlapping & Accuracy & Full & 0.911 & 0.0217 \\
  Overlapping & Accuracy & Vanilla & 0.916 & 0.00236 \\
  Overlapping & F1-score & Full & 0.911 & 0.0219 \\
  Overlapping & F1-score & Vanilla & 0.916 & 0.00236 \\\bottomrule

@@

\subsubsection{Plausibility}\label{plausibility-3}

@@

\subsubsection{Cost}\label{cost-1}

@@

\subsection{Other Parameters}\label{sec-app-grid-train}

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-train-final-run-train}: Training Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]

@@

\subsubsection{Accuracy}\label{accuracy-2}

@@

  \toprule
  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
  \endfirsthead
  \toprule
  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
  \endhead
  \bottomrule
  \multicolumn{5}{r}{Continuing table below.}\\
  \bottomrule
  \endfoot
  \endlastfoot
  Circles & Accuracy & Full & 0.995 & 0.00431 \\
  Circles & Accuracy & Vanilla & 0.998 & 0.000566 \\
  Circles & F1-score & Full & 0.995 & 0.00432 \\
  Circles & F1-score & Vanilla & 0.998 & 0.000566 \\
  Linearly Separable & Accuracy & Full & 0.999 & 0.00231 \\
  Linearly Separable & Accuracy & Vanilla & 1 & 0 \\
  Linearly Separable & F1-score & Full & 0.999 & 0.00231 \\
  Linearly Separable & F1-score & Vanilla & 1 & 0 \\
  Moons & Accuracy & Full & 0.996 & 0.0136 \\
  Moons & Accuracy & Vanilla & 0.988 & 0.022 \\
  Moons & F1-score & Full & 0.996 & 0.0136 \\
  Moons & F1-score & Vanilla & 0.988 & 0.022 \\
  Overlapping & Accuracy & Full & 0.914 & 0.00563 \\
  Overlapping & Accuracy & Vanilla & 0.918 & 0.00116 \\
  Overlapping & F1-score & Full & 0.914 & 0.0057 \\
  Overlapping & F1-score & Vanilla & 0.918 & 0.00116 \\\bottomrule

@@

\subsubsection{Plausibility}\label{plausibility-4}

@@

\subsubsection{Cost}\label{cost-2}

@@

\section{Tuning Key Parameters}\label{sec-app-tune}

@@

Finally, to increase the proportion of mature counterfactuals for some
datasets, we have also investigated the effect on the learning rate
\(\eta\) for the counterfactual search and even smaller regularization
strengths for a fixed decision threshold of 0.5
(Figure~\ref{fig-tune_lr-plaus-adult} to
Figure~\ref{fig-tune_lr-plaus-over}). For the given low decision
threshold, we find that the learning rate has no discernable impact on
the proportion of mature counterfactuals
(Figure~\ref{fig-tune_lr-mat-adult} to
Figure~\ref{fig-tune_lr-mat-over}). We do notice, however, that the
results for MNIST are much improved when using a low value
\(\lambda_{\text{reg}}\), the strength for the engery regularization:
plausibility is increased by up to \textasciitilde10\%
(Figure~\ref{fig-tune_lr-plaus-mnist}) and the proportion of mature
counterfactuals reaches 100\%.

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Package Version (Reproducibility)}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-warning-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-warning-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]

@@

\subsection{Key Parameters}\label{sec-app-tune-key}

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-tune-train}: Training Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]

@@

\subsubsection{Plausibility}\label{plausibility-5}

@@

\subsubsection{Proportion of Mature CE}\label{proportion-of-mature-ce}

@@

\subsection{Learning Rate}\label{sec-app-tune-lr}

@@

\subsubsection{Plausibility}\label{plausibility-6}

@@

\subsubsection{Proportion of Mature CE}\label{proportion-of-mature-ce-1}
